---
title: "Activity on weekdays and weekends"
author: "Jameaba"
date: "October 18, 2015"
output: html_document: 
    number_sections: yes
    toc: yes
---
    
    #An Analysis of Data on Activity 
    
    **Reading data into R Studio**
    
    ```{r, echo=TRUE}
    stappes = read.table("activity.csv", header = T, sep =",")
    stappdf = data.frame(stappes)
    stappdf1 <- na.omit(stappdf)
    ```
    *A summary of variables in the activity dataset*
    
    ```{r, echo=TRUE}
    summary(stappes)
    ```
    ##Reading file into R Studio

    ```{r, echo= TRUE}
    stappes = read.table("activity.csv", header = T, sep =",")
    ```

    *Creating a dataframe of the activity dataset*
    
    ```{r, echo=TRUE}
    stappdf = data.frame(stappes)
    ```

    **Total number of steps taken per day and naming the categorized       variables**
    
    ```{r, echo=TRUE}
    stappes2 <- aggregate.data.frame(stappes$steps,
    by = list(stappes$date), FUN = sum)
    names(stappes) <- c("date", "sumsteps")
    ```

    **Make a histogram of the total number of steps taken each day**
    
    ```{r, echo=TRUE, fig.width=7, fig.height=5, echo=TRUE}
    windows()
    hist(stappes2$steps) 
    ```
    *Alternatively, histogram can be constructed as follows*
    
    ```{r, fig.width=7, fig.height=5,echo=TRUE}
    hist(stappes2$steps,xlab="Sum steps per day", type ="1", col             ="red", main = "sum steps per day")
    ```
    *Calculate average  steps by day*
    
    ##Average daily activity pattern
    
    stappes1 <- aggregate(stappes$steps,
    by =  list(stappes$date), FUN = mean)
    names(stappes1) <- c("date", "meansteps")
    print(stappes1)
     
     ```{r qplot, fig.width=7, fig.height=7, echo=TRUE}
    windows()
    library (ggplot2)
    qplot(stappes$interval, stappes$steps,xlab ="mean interval             per day", ylab= "mean steps per day", color= "wheat",  type = "1",      main = "Five-minute interval by  steps",  geom = point",   
    "smooth", method =  "lm" )
    ```

    *The interval that has the largest number of steps occurs on            2012-10-06* 
    
    #Imputing missing values
    
    *Calculating total number of missing values*
    
    
    ```{r, echo=TRUE}
    stappes_msvas=sapply(stappes, function(x) sum(is.na(x)))
    stappes_msvas
    ```
    Reported number of  mssing values is **2304** for steps variable 
    #Filling in the missing values using the mean


    ```{r,echo=TRUE}
    stappes$steps[is.na(stappes$steps)]
    mean(stappes$steps, na.rm = TRUE)
    stappes$steps[is.na(stappes$steps)] <- mean(stappes$steps,             na.rm = TRUE)
    stat<-data.frame(stappes$steps)
    ```

    *Missing values for interval*
    
    ```{r, echo=TRUE}
    stappes$interval[is.na(stappes$interval)]
    median(stappes$interval, na.rm = TRUE)
    stappes$interval[is.na(stappes$interval)]<-                             median(stappes$interval, na.rm = TRUE)

    stappes$interval
    ```

    # data frame with MAs steps and interval variables filled

    ```{r, echo=TRUE}
    stappes_ma_filled <- data.frame(stappes$steps, stappes$date,           stappes$interval)
    print(stappes_ma_filled)
    ```

    *Histogram of the steps taken by day with MA filled*
    
    ```{r plot,fig.width=7, fig.height=5, echo=TRUE}
    windows())
    hist(stappes_ma_filled$stappes.steps)
    ```

    *The mean and medium of total steps by day
    **Mean of steps is 37.38; median is 0** [unchanged]

    ```{r, echo=TRUE}
    summary(stappes_ma_filled) 
    ```
    *The effect of filling in inmissing values with means seems to          enhance smoothness of the   distribution*.

    ```{r, echo =TRUE}
    stappdf = data.frame(stappes)
    df = transform(stappdf, steps = ifelse(is.na(steps), mean(steps,       na.rm=TRUE), steps))
    df1 =transform(stappdf, interval = ifelse(is.na(interval),         
    mean(interval, na.rm=TRUE), interval)) 
    ```
    *Graphically*
    
    ```{r qplot,  fig.width=7, fig.height=5, echo=TRUE}
    Wndows()
    library(ggplot2)
    qplot(interval, stepp1, data = stap, main = " Five_minute interval     by  steps", geom  =  c("point","smooth"),  method ="lm")
    ```
    ##Activity patterns on weekends and weekdays

    *Converting dates into weekends and weekdays*
    
    ```{r, echo=TRUE}
    mydates <- as.Date(stappeplete$stappes.date)
    mydadates1 <-weekdays(mydates)
    ```
    *creating 'differentiators', the two levels and labels*
    
    ```{r, echo=TRUE}
    mydates2<- c("Monday", "Tuesday", "Wednesday", "Thursday","Friday")
    
    mydates3 <- factor((weekdays(mydates) %in% mydates2), 
    levels=c(FALSE, TRUE), labels = c("Weekend",  "weekday"))
    ```
    *Creation of dataframe with a new factor variable*
    
    
    ```{r, echo = TRUE}
    stappeplete1<-data.frame(stappeplete, mydate4$date_by_day)
    names(stappeplete1) <- c("steps",  "date",  "interval",   
    "day_week")
    ```
    *Making things abit* 
    
    ```{r, echo=TRUE}
    stapt <- data.frame(stappeplete1)
    ```
    *Creating subsets of weekend days and weekday days*
    
    ```{r, echo= TRUE}
    
    (stapt2 <- stapt[stapt$day_week == "weekend", ])
    
    (stapt3 <- stapt[stapt$day_week == "weekday", ])
    ```

    # Plot intervals by weekends and weekdays

    ```{r, fig.width=7, fig.height=5, echo=TRUE}
    windows()
    par(mfrow = c (2,2))
    plot(stapt2$interval, stapt2$steps,type="l", xlab   ="Interval",       ylab ="Steps on weekends", col = "blue")
    plot(stapt3$interval, stapt3$steps, type="l", xlab  = "Interval",      ylab="Steps on weekdays", col ="magenta" )
    ```
    *There is apparently a marked difference in the pattern of activity     on weekdays and weekends. On weekdays actvity is very intensive  
    during early mmorning, peaks somewhat quickly and starts to fall 
    off, albeit with intermitent spurts along the way. For the weekends     activity seems intense during later hours. It is also important to
    note that the concentration of pockets of time of intense activity
    seems to occur more in batches with respect to weekends than on  
    weekdays. Another important difference between the two categories 
    lies in the high concentration of activity (more compact) compared
    with somewhat slightly paced activity on weekends.
    On weekdays and weekends. On weekdays activity is very intensive 
    during early morning, peaks somewhat quickly and starts to fall
    off, albeit with intermitent spurts along the way. For the weekends     activity seems intense during later hours. It is also important to     note that the concentration of pockets of time of intense activity 
    seems to occur more in batches with respect to weekends than on  
    weekdays. Another important difference between the two categories  
    lies in the high concentration of activity (more compact) compared 
    with somewhat slightly paced activity on weekends.*
